\LARGE \bf Artifical Cognition for Social Human-Robot Interaction:\\ An Implementation

### 2016-06-16 paper.tex
Taken into account most of the reviewer 2 comments

\usepackage[nomargin, marginclue, footnote]{fixme}

@@

Human-robot interaction is a challenge for Artificial Intelligence. This field
lays at the crossroad of several domains of AI and requires to tackle them in a
holistic manner: modelling humans and human cognition; acquiring, representing,
manipulating in a tractable way abstract knowledge at the human level; reasoning
on this knowledge to make decisions; and eventually instantiating those
decisions into physical actions both legible to and in coordination with humans.
Many AI techniques are invited, from visual processing to symbolic reasoning,
from task planning to \emph{theory of mind} building, from reactive control to
action recognition and learning.

@@

The robot controller (with the help of a task planner) decides what action to
execute next, and who should perform it from the robot or the human (or both in
case of a joint action), and finally controls or monitors its execution. The
operation continues until the goal is achieved, is declared unachievable or is
abandoned by the human~\cite{Klein2004}.

@@

Besides, such abilities should be designed and implemented in a yask-independent
manner, and should provide sufficient levels of parametrization such as they adapt to
various environments, different tasks and variable levels of engagement of the
robot, ranging from teammate behaviour to assistant or pro-active helper.

@@

The remaining of the article discusses the robotic architecture that we have built to
tackle the AI challenges of human-robot interaction, and how its relates to
other approaches. We organise this discussion in four sections.

@@

In our architecture, knowledge manipulation relies on a central server (the {\sc
Oro} server~\cite{Lemaignan2010}, Figure~\ref{fig|spark-oro} top) which stores
knowledge as it is produced by each of the other deliberative components (the
clients). It exposes a {\tt json}-based RPC API to query the knowledge
base~\cite{lemaignan2012kbapi}.  We represent knowledge as RDF triples in the
OWL sub-language~\footnote{\url{http://www.w3.org/TR/owl2-overview/}}. Every
time triples are added or removed from the knowledge base, a Description Logics
reasoner ({\sc Pellet}\footnote{\url{https://github.com/complexible/pellet}})
classifies the whole ontology and inserts all possible inferred triples.  The
clients of the {\sc Oro} server are in charge of managing themselves the
knowledge (when to add, when to update, when to retract knowledge): no
meta-semantics are carried over that would let the server manage these dynamics
itself.\footnote{With one exception: so-called \emph{memory profiles} let
the server automatically discard facts (\ie forget about them) after a specific
period of time. We present this feature at section~\ref{memory}.}

@@

Contrary to similar projects like KnowRob~\cite{Tenorth2009a} that relies on
the concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
needed, we have an \emph{explicit} approach where we greedily compute and assert
symbolic statements (like spatial relations between objects, see
Section~\ref{sect|sit-ass}). This design choice trades scalability for explicit
reasoning: at any time, the full belief state is explicitly stated, and
provides the reasoner with the largest possible inference domain. This is of
special importance for an \emph{event-based} architecture like ours (see
Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
we are ``looking for'', and we need instead to greedily infer as much as possible
to build a belief state as comprehensive as possible.

@@

Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
of the ontology is generally difficult: it pertains to abstract concepts whose
mutual relations comes to philosophical debates. The upper taxonomy of {\sc
OpenCyc} represents a relative consensus, at least within the semantic Web
community. Then, because it is a well established project with numerous links to
other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
OpenCyc} concepts ensures that the knowledge stored by the robot can be shared
or extended with well-defined semantics. The concept of \emph{Object} represents
a typical case of ambiguous meaning: in everyday conversation, an object is a
relatively small physical thing, that can be typically manipulated. A human is
not usually considered as an object. {\sc OpenCyc} however precisely defines an
object as anything at least \emph{partially tangible}. This includes obviously
the humans, and actually many other entities that would not be commonly said to
be objects (the Earth for instance). Thus the importance of relying on
well-defined and standard semantics to exchange information between artificial
systems in a non-ambiguous way.

@@

Lastly, the {\sc Oro} common-sense ontology contains several rules and class
expressions that encode non-trivial inferences. The definition of
\concept{Bottle} as found in the {\sc Oro} ontology is a typical example:
$$
\text{\tt Bottle} \equiv \text{\tt Container} \wedge 
                         \text{\tt Tableware} \wedge
                         \text{\tt hasShape} \in \text{\tt CylinderShape}^{\mathcal{I}} \wedge
                         \text{\tt hasMainDimension} \in [0.1, 0.3]
$$

@@

\emph{Symbol grounding} connects hence the knowledge model to the perception and
actuation capabilities of the robot, and our architecture adopts here also a
bidirectional approach: \emph{bottom-up} and \emph{top-down}.  The different
components that we have mentioned so far provide a bottom-up grounding process:
geometric reasoning and dialogue processing modules constantly build and push
new symbolic contents about the world to the knowledge base where it becomes
accessible to decisional layers. In parallel, the knowledge base relies on
reasoning in a top-down way to produce new facts, \ie new beliefs that may lead
to actual decision making and physical actions.

@@

{\sc Spark} runs and continuously updates the knowledge base at about 10Hz. At each step, it re-computes spatial
relations and affordances for the whole scene, and send the delta (new
relations and relations that do not hold anymore) to the knowledge base. This
approach may raise scalability concerns (we did not however hit major
performance issues in our constrained scenarios), but simplifies the
management of the \emph{dynamics} of the knowledge (\emph{When do I discard
outdated knowledge?  When do I update it?}). Since it is equivalent to a reset
of the reasoner domain, it also hides issues linked to non-monotonic reasoning
(see~\cite{McCarthy2007} for a discussion on that question).

@@

Humans commonly refer to the positions of objects with symbolic descriptors
(like \emph{on}, \emph{next to}...) instead of precise, absolute positions.
These type of descriptors have been studied in the context of language grounding
\cite{O'Keefe1999,Matuszek2010,Regier2001,Kelleher2006,Blisard2005}.  {\sc
Spark} distinguishes between agent-independent symbolic locations (allocentric
spatial relations) and agent-dependent, relative locations (egocentric spatial
relations).

{\sc Spark} computes three main agent-independent relations based on the
bounding box and centre of mass of the objects
(Figure~\ref{fig|spark-oro}\emph{a}): \concept{isOn} holds when an object $O_1$
is on another object $O_2$, and is computed by evaluating the centre of mass of
$O_1$ according to the bounding box of $O_2$.  \concept{isIn} evaluates if an
object $O_1$ is inside another object $O_2$ based on their bounding boxes
$BB_{O_1}$ and $BB_{O_2}$.  \concept{isNextTo} indicates whether an object $O_1$
is next to another object $O_2$. We do not use a simple distance threshold to
determine if two objects are next to each other since the relation is highly
dependent on the dimensions of the objects. For instance, the maximum distance
between large objects (\eg two houses) to consider them as being next to each
other is much larger than the maximum distance we would consider for two small
objects (\eg two bottles). Thus, the distance threshold is scaled with the
objects' size. {\sc Spark} also compute symbolic facts related to agent
independent world dynamics.  The predicate \concept{isMoving} states, for each
tracked entity, whether it is currently moving or not.

@@

Monitoring human activity is needed by the execution controllers to track the
engagement of the human and the progress of his/her actions. Full human action
and activity recognition is a task that requires knowledge and reasoning both on
high-level facts like goals, intentions and plans, as well as bottom-up data
from agent and object motions. {\sc Spark} implements a set of simple temporal and
geometric heuristics on human hand trajectories and possible objects placements
to recognize simple atomic actions. Those primitive actions are assessed through
monitoring situations like ``an empty hand is close to an object on a table''
(precursor for a \emph{pick}), or ``a hand holding an object is over a
container'' (precursor for a \emph{throw}).  {\sc Spark} recognises a set of
such primitives that are relied upon by the execution controller to track the
state transitions in the human plans.

@@

We acquire natural speech input from the human participants through a custom
Android-based interface that relies on the Google speech recognition API for
speech-to-text and relays the textual transcript to the robot. It is parsed into
a grammatical structure (\emph{Part of Speech} tagging), and the atoms of each
sentence are resolved with the help
of the ontology to ground concepts like objects (\ie when a user says ``pick the
can'', it resolves to which instance of \emph{can} the user is referring to) and
actions.  Sentences are classified into questions, desires and statements, and
processed accordingly.  Figure~\ref{dialogs|ex} gives an example of the
processing of a simple, non-ambiguous command.

@@

        \emph{Generated query} \\
        \hline
        \concept{find(?obj type Book)} \\ 
        \hspace{0.2cm}$\Rightarrow$ \concept{?obj = book\_1} \\
        ~\\
        ~\\
        ~\\
    \end{tabular}
    \begin{tabular}{l}

        \emph{Newly created statements}\\
        \hline
        \stmt{human\_1 desires sit\_1} \\
         \stmt{sit\_1 type Give} \\
         \stmt{sit\_1 performedBy myself} \\
         \stmt{sit\_1 actsOnObject book\_01} \\
         \stmt{sit\_1 receivedBy human} \\
    \end{tabular}

    \caption{Processing of a simple, non-ambiguous command, taken
        from~\cite{Lemaignan2011a}. Thematic roles ({\tt performedBy}, {\tt
        actsOnObject}, {\tt receivedBy}) are automatically extracted by 
    {\sc Dialogs} from the imperative sentence ``Give me the book.''.}

@@

The second level of integration of multi-modality is implicit. By continuously
computing symbolic properties from the geometric model, richer descriptions and
hence discrimination possibilities are available. For instance, the robot may
compute that one bottle is next to a glass while another one stands alone, and
these symbolic descriptions are transparently re-used in a dialogue context to
generate unambiguous references to discriminate between similar objects: ``do
you mean the bottle that is next to the glass?'' (\cite{Ros2010b} provides a
detailed account of our approach to interactive concept discrimination, along
with the related algorithms).

@@

The HATP planning framework extends the traditional HTN (Hierarchical Task
Network) planning domain representation and semantics by making them more
suitable to elaborate plans which involve humans and robots acting together
toward a joint goal. HATP is used by the robot to elaborate human-robot
\emph{shared plans}~\cite{Grosz1996,Clark1996,Kemp2007} which are then used to
anticipate human action, to propose to human to act or even to ask help from
the human.

@@

HATP includes mechanisms called \emph{social rules} to filter plans so as to
keep only those considered as suitable for human-robot interaction. The planner
allows the specification of the following filtering criteria:

@@

\emph{Control of intricacy}: avoids plans with too many interdependencies
between the actions of agents mentioned in the plan, as a problem with
executing just one of those actions could invalidate the entire
plan. Also intricate human-robot activity may cause discomfort since
the human will find himself repeatedly in a situation where his is
waiting for the robot to act,

@@

One of the main role of {\sc Shary} is to control the production and execution
of shared plans. This means essentially context-based refinement and robot
actions execution, as well as monitoring of those achieved by its human
partner. One of the key design goal is to build such abilities in a generic
way, and to provide several levels of parametrisation allowing to adapt to
various environments, and various levels of involvement of the robot ranging
from teammate behaviour to assistant or proactive helper. Based on this,
the robot controller invokes the adequate human-aware planners and react to
events triggered by the {\sc Oro} knowledge base (as described below).

{\sc Shary}'s originality, as an execution control system, lies in its ability
to take into account not only the task achievement but also the communication
and monitoring needed to support interactive task
achievement~\cite{Rich1997,Sidner2005} in a flexible manner. {\sc Shary} allows
to bind action execution to \emph{communication policies} in order to produce
signals towards the human and to react to human actions. The
\emph{communication act} is the central concept in this formalism. It
represents an information exchange between the two agents and plays the role of
a transition condition. This exchange can be realised through dialogue, by an
expressive motion or a combination of the two. It enables each agent to
communicate their beliefs about the task to be achieved, in order to share
mutual knowledge and to synchronise their activities.  This is done through
real-time task-based situation assessment achieved by the combination of {\sc
Shary} monitoring processes and {\sc Oro} inference
mechanisms~\cite{fiore2014}.

\subsubsection{Event-Driven Control}

@@

This allows us to write reactive robot controllers with a high level of
expressiveness: for instance, by subscribing to the event \stmt{human1 desires
?action, ?action type Take, ?action actsOnObject ?obj, ?obj type Toy}, we could trigger a
behaviour when the human expresses (through dialogue, gestures...) that he
wants a toy.

@@

\subsection{\emph{Cleaning the table} Study}

@@

This study (Figure~\ref{fig|cleantable-video}) exhibits a richer decision-making
process. The {\sc Oro} server is used in conjunction with the HATP symbolic task
planner and the {\sc Shary} execution controller to produce and execute a shared
plan.

@@

Robotics is traditionally regarded as the prototypical instance of
\emph{embodied} artificial intelligence, and this dimension is especially
prevalent in human-robot interaction, where agents have to share and
collaborate in a joint physical environment.  This leads to a tight coupling
between the symbolic and the geometric realms: while AI at its origins was
mostly concerned with symbolic models, it has been since recognised that not
only the mind is not a purely abstract system, disconnected from the physical
world, but even more, cognition fundamentally relies on its relation to the
physical world (so-called \emph{embodied cognition}). Varela~\cite{Varela1992}
is one of the main discoverers of these mechanisms, and coined the concept of
\emph{enactivism} as the theoretical framework that study the links between
cognition, embodiment and actions.  This has since been thoroughly studied in
robotics and artificial intelligence (Pfeifer and
Bongard~\cite{pfeifer2007body} is one of the reference).

The challenge of symbol grounding is also tightly linked to this issue. It
corresponds to the identification or creation, and then, maintenance of a link
between the symbol (the syntactic form of knowledge that the computer
manipulates) and its semantics (its meaning, anchored in the world). The
relation between the symbol, the referent of the symbol, and mediating minds
is classically referred as the \emph{semantic triangle} and has been
extensively studied in linguistics. The issue of grounding is well known in
cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
question: ``how the semantic interpretation of a formal symbol system can be
made intrinsic to the system?''. This issue has a major practical importance in
robotic: for a robot to be both endowed with a symbolic representational and
reasoning system, and able to \emph{act} in the physical world, it must ground
its knowledge.

@@

Our context, where a robot has to achieve a task together with
humans, raises issues to be tackled by the robots' decisional components, and
that we can summarises as ``W-questions'': \emph{What}, \emph{Who}, \emph{When},
\emph{Where} and \emph{How}?

\emph{What to do next}, at different levels of abstractions and while taking
into account not only the current state but also the long term goals, is the
basic question for an intelligent robot. Its is made more complex here by having
to deal with the partially observable physical and mental state of the human
partner, and by the extended set of possible actions.

@@

@@

Rachid -- TDB: what are the other 'deliberative architecture for socially interactive
robots' and where/how ours is better?

Discussion on deliberative architectures != cognitive architecture: a \emph{cognitive
architecture} is usually understood as an artificial yet principled model of (human)
cognition, while we propose a practical approach to consistently combine a large
set of independent cognitive processes by bridging them through high-level semantics.

>>>>

\usepackage[draft,nomargin, marginclue, footnote]{fixme}

@@

Human-Robot Interaction (HRI) represents a challenge for Artificial Intelligence
(AI). It lays at the crossroad of many subdomains of AI and in effect, it calls
for their integration: modelling humans and human cognition; acquiring,
representing, manipulating in a tractable way abstract knowledge at the human
level; reasoning on this knowledge to make decisions; eventually instantiating
those decisions into physical actions both legible to and in coordination with
humans.  Many AI techniques are mandated, from visual processing to symbolic
reasoning, from task planning to \emph{theory of mind} building, from reactive
control to action recognition and learning.

@@

The robot controller (with the help of a task planner) decides what action to
execute next, and who should perform it, from the robot or the human (or both in
case of a joint action). It finally controls and monitors its execution. The
operation continues until the goal is achieved, is declared unachievable or is
abandoned by the human~\cite{Klein2004}.

@@

Besides, such abilities should be designed and implemented in a yask-independent
manner, and should provide sufficient levels of parametrization, so that they adapt to
various environments, different tasks and variable levels of engagement of the
robot, ranging from teammate behaviour to assistant or pro-active helper.

@@

The remaining of the article discusses the robotic architecture that we have
built to approach social and autonomous human-robot interaction, and how it
relates to other approaches. We organise this discussion in four sections.

@@

In our architecture, knowledge manipulation relies on a central server (the {\sc
Oro} server~\cite{Lemaignan2010}, Figure~\ref{fig|spark-oro} top) which stores
knowledge as it is produced by each of the other deliberative components (the
clients). It exposes a {\tt json}-based RPC API to query the knowledge
base~\cite{lemaignan2012kbapi}.  We represent knowledge as RDF triples in the
OWL sub-language~\footnote{\url{http://www.w3.org/TR/owl2-overview/}}. Every
time triples are added or removed from the knowledge base, a Description Logics
reasoner ({\sc Pellet}\footnote{\url{https://github.com/complexible/pellet}})
classifies the whole ontology and inserts all possible inferred triples.  The
clients of the {\sc Oro} server are in charge of managing themselves the
knowledge (when to add, when to update, when to retract knowledge): no
meta-semantics are carried over that would let the server manage itself these
dynamics.\footnote{With one exception: so-called \emph{memory profiles} let the
server automatically discard facts (\ie forget about them) after a specific
period of time. We present this feature at section~\ref{memory}.}

@@

Contrary to similar projects like KnowRob~\cite{Tenorth2009a} that relies on the
concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
needed, we have an \emph{explicit} approach where we greedily compute and assert
symbolic statements (like spatial relations between objects, see
Section~\ref{sect|sit-ass}). For instance, whenever a client queries KnowRob to
know if \stmt{object1 isOn object2}, KnowRob calls a geometric reasonner to
evaluate if this specific spatial relation holds at that point in time. With our
approach, spatial relations are instead computed and asserted {\it a priori}.
This design choice trades scalability for explicit reasoning: at any time, the
full belief state is explicited (which may be computationally expensive), and
therefore provides the reasoner with the largest possible inference domain. This
is of special importance for an \emph{event-based} architecture like ours (see
Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
we are ``looking for'', and we need instead to greedily infer as much as
possible to build a belief state as comprehensive as possible.

@@

Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
of the ontology is generally difficult: it pertains to abstract concepts whose
mutual relations comes to philosophical debates. The upper taxonomy of {\sc
OpenCyc} represents a relative consensus, at least within the semantic Web
community. Then, because it is a well established project with numerous links to
other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
OpenCyc} concepts ensures that the knowledge stored by the robot can be shared
or extended with well-defined semantics. The concept of \emph{Object} represents
a typical case of ambiguous meaning: in everyday conversation, an object is a
relatively small physical thing, that can be typically manipulated. A human is
not usually considered as an object. {\sc OpenCyc} however precisely defines an
object as anything at least \emph{partially tangible}. This includes obviously
the humans, and actually many other entities that would not be commonly said to
be objects (the Earth for instance).\footnote{In such cases of discrepancy
between {\sc OpenCyc} concepts and human terminology, we manually label the {\sc
OpenCyc} concepts with appropriate human names: for instance, the OpenRobots
Ontology associates the label ``\emph{object}'' to the concept
\concept{cyc:Artifact} instead of the concept \concept{cyc:PartiallyTangible}.
These labels are used in priority during the grounding of verbal interactions.}
Thus the importance of relying on well-defined and standard semantics to
exchange information between artificial systems in a non-ambiguous way.

@@

Lastly, the {\sc Oro} common-sense ontology contains several rules and class
expressions that encode non-trivial inferences. The definition of
\concept{Bottle} as found in the {\sc Oro} ontology is a typical example:
$$
\text{\tt Bottle} \equiv \text{\tt Container} \wedge 
                         \text{\tt GraspableObject} \wedge
                         \text{\tt hasShape} \in \text{\tt CylinderShape}^{\mathcal{I}} \wedge
                         \text{\tt hasMainDimension} \in [0.1, 0.3]
$$

@@

\emph{Symbol grounding} connects hence the knowledge model to the perception and
actuation capabilities of the robot. The different components that we have
mentioned so far provide a bottom-up grounding process: geometric reasoning and
dialogue processing modules constantly build and push new symbolic contents
about the world to the knowledge base where it becomes accessible to decisional
layers. We detail in the next sections this process.

@@

{\sc Spark} runs and continuously updates the knowledge base at about 10Hz. At
each step, it re-computes spatial relations and affordances for the whole scene,
and send the delta (new relations and relations that do not hold anymore) to the
knowledge base. This approach may raise scalability concerns (we did not however
hit major performance issues in our constrained scenarios) as well as prevent
reasoning on the situation history, but simplifies the management of the
dynamics of the knowledge (\emph{When do I discard outdated knowledge?  When do
I update it?}). Since it is equivalent to a reset of the reasoner domain, it
also essentially nullifies issues linked to non-monotonic reasoning
(see~\cite{McCarthy2007} for a discussion on that question).

@@

Humans commonly refer to the positions of objects with symbolic descriptors
(like \emph{on}, \emph{next to}...) instead of precise, absolute positions
(qualitative spatial reasoning). These type of descriptors have been studied in the context of language grounding
\cite{O'Keefe1999,Matuszek2010,Regier2001,Kelleher2006,Blisard2005}.  {\sc
Spark} distinguishes between agent-independent symbolic locations (allocentric
spatial relations) and agent-dependent, relative locations (egocentric spatial
relations).

{\sc Spark} computes three main agent-independent relations~\cite{Sisbot2011}
based on the bounding box and centre of mass of the objects
(Figure~\ref{fig|spark-oro}\emph{a}): \concept{isOn} holds when an object $O_1$
is on another object $O_2$, and is computed by evaluating the centre of mass of
$O_1$ according to the bounding box of $O_2$.  \concept{isIn} evaluates if an
object $O_1$ is inside another object $O_2$ based on their bounding boxes
$BB_{O_1}$ and $BB_{O_2}$.  \concept{isNextTo} indicates whether an object $O_1$
is next to another object $O_2$. We do not use a simple distance threshold to
determine if two objects are next to each other since the relation is highly
dependent on the dimensions of the objects. For instance, the maximum distance
between large objects (\eg two houses) to consider them as being next to each
other is much larger than the maximum distance we would consider for two small
objects (\eg two bottles). Thus, the distance threshold is scaled with the
objects' size. {\sc Spark} also compute symbolic facts related to agent
independent world dynamics.  The predicate \concept{isMoving} states, for each
tracked entity, whether it is currently moving or not.

@@

Monitoring human activity is needed by the execution controllers to track the
engagement of the human and the progress of his/her actions. Full human action
and activity recognition is a task that requires knowledge and reasoning both on
high-level facts like goals, intentions and plans, as well as bottom-up data
from agent and object motions. {\sc Spark} implements a set of simple temporal and
geometric heuristics on human hand trajectories and possible objects placements
to recognize simple atomic actions. Those primitive actions are assessed through
monitoring situations like ``an empty hand is close to an object on a table''
(precursor for a \emph{pick}), or ``a hand holding an object is over a
container'' (precursor for a \emph{throw}).  {\sc Spark} recognises a set of
such primitives. When combined with the other geometric computations and a predictive
plan of the human actions (see Section~\ref{hatp}), the execution controller can
track the fulfilment of the pre- and post-conditions of the predicted human
actions. As a result, the robot can monitor the engagement of the human and the
overall progress of the human-robot shared plan.

@@

We acquire natural speech input from the human participants through a custom
Android-based interface that relies on the Google speech recognition API for
speech-to-text and relays the textual transcript to the robot. It is parsed into
a grammatical structure (\emph{Part of Speech} tagging), and the atoms of each
sentence are resolved with the help of the ontology to ground concepts like
objects (\ie when a user says ``pick the can'', it resolves to which instance of
\emph{can} the user is referring to) and actions.  Figure~\ref{dialogs|ex} gives
an example of the processing of a simple, non-ambiguous command, while the first
study (Section~\ref{moving-london}) walks through more complex examples.
Heuristics (like the presence of a question mark or the use of imperative mood)
are used to classify the sentences into questions, desires or statements, and
{\sc Dialogs} processes them accordingly (by answering questions or
updating the knowledge base).

@@

        \emph{Generated query to ontology} \\
        \hline
        \concept{find(?obj type Book)} \\ 
        \hspace{0.2cm}$\Rightarrow$ \concept{?obj = book\_1} \\
        ~\\
        ~\\
        ~\\
    \end{tabular}
    \begin{tabular}{l}

        \emph{Newly created statements}\\
        \hline
        \stmt{human\_1 desires sit\_1} \\
         \stmt{sit\_1 type Give} \\
         \stmt{sit\_1 performedBy myself} \\
         \stmt{sit\_1 actsOnObject book\_01} \\
         \stmt{sit\_1 receivedBy human\_1} \\
    \end{tabular}

    \caption{Processing of a simple, non-ambiguous command, taken
        from~\cite{Lemaignan2011a}. Thematic roles ({\tt performedBy}, {\tt
        actsOnObject}, {\tt receivedBy}) are automatically extracted by 
    {\sc Dialogs} from the imperative sentence ``Give me the book.'' The new
    resulting statements (right column) are added to the knowledge base, and
    may eventually trigger an event in the execution controller (see
    Section~\ref{events}).}

@@

The second level of integration of multi-modality is implicit. By continuously
computing symbolic properties from the geometric model, richer descriptions and
hence discrimination possibilities are available. For instance, the robot may
compute that one bottle is next to a glass while another one stands alone, and
these symbolic descriptions are transparently re-used in a dialogue context to
generate unambiguous references to discriminate between similar objects: ``do
you mean the bottle that is next to the glass?''. \cite{Ros2010b} provides a
detailed account of our approach to interactive concept clarification and
discrimination, along with the related algorithms.

@@

The HATP planning framework extends the traditional HTN (Hierarchical Task
Network) planning domain representation and semantics by making them more
suitable to elaborate plans which involve humans and robots acting together
toward a joint goal. HATP is used by the robot to elaborate human-robot
\emph{shared plans}~\cite{Grosz1996,Clark1996,Kemp2007} which are then used to
anticipate human action, to suggest a course of action to humans, or even to ask
help from the human.

@@

HATP includes mechanisms called \emph{social rules} to filter plans so as to
keep only those considered as suitable for human-robot interaction. The planner
allows for the specification of the following filtering criteria:

@@

\emph{Simplicity}: avoids plans with too many interdependencies
between the actions of agents mentioned in the plan, as a problem with
executing just one of those actions could invalidate the entire
plan. Also intricate human-robot activity may cause discomfort since
the human will find himself repeatedly in a situation where his is
waiting for the robot to act,

@@

One of the main roles of {\sc Shary} is to control the production and execution
of shared plans. This means essentially context-based refinement and robot
actions execution, as well as monitoring of those achieved by its human
partner. One of the key design goal is to build such abilities in a generic
way, and to provide several levels of parametrisation allowing to adapt to
various environments, and various levels of involvement of the robot ranging
from teammate behaviour to assistant or proactive helper. Based on this,
the robot controller invokes the adequate human-aware planners and react to
events triggered by the {\sc Oro} knowledge base (as described below).

{\sc Shary}'s originality, as an execution control system, lies in its ability
to take into account not only the task achievement but also the communication
and monitoring needed to support interactive task
achievement~\cite{Rich1997,Sidner2005} in a flexible manner. {\sc Shary} allows
to bind action execution to \emph{communication policies} in order to produce
signals towards the human and to react to human actions. The
\emph{communication act} is the central concept in this formalism. It
represents an information exchange between the two agents and plays the role of
a transition condition. This exchange can be realised through dialogue, by an
expressive motion or a combination of the two. It enables each agent to
communicate their beliefs about the task to be achieved, in order to share
mutual knowledge and to synchronise their activities.  This is done through
real-time task-based situation assessment achieved by the combination of {\sc
Shary} monitoring processes and {\sc Oro} inference
mechanisms~\cite{fiore2014}. \fixme{Rachid, reviewer 2 asks for more details
about this process -> M. Fiore work?}

\subsubsection{Event-Driven Control}
\label{events}

@@

This allows us to write reactive robot controllers with a high level of
expressiveness: for instance, by subscribing to the event \stmt{human1 desires
?action, ?action type Give, ?action actsOnObject ?obj, ?obj type Book}, we could
trigger a behaviour when the human expresses (through dialogue, gestures...)
that he wants the robot to give her a book.

@@

\subsection{\emph{Cleaning the table} Study}
\label{sec:cleantable}

@@

This second study demonstrates a richer decision-making process where the {\sc
Oro} server is used in conjunction with the HATP symbolic task planner and the
{\sc Shary} execution controller to produce and execute a shared plan. The task
consists in cooperatively cleaning a table by moving objects into their target
bins (Figure~\ref{fig|cleantable-video}).

@@

Robotics is traditionally regarded as the prototypical instance of
\emph{embodied} artificial intelligence, and this dimension is especially
prevalent in human-robot interaction, where agents have to share and
collaborate in a joint physical environment.  This leads to a tight coupling
between the symbolic and the geometric realms: while AI at its origins was
mostly concerned with symbolic models, it has been since recognised that not
only the mind is not a purely abstract system, disconnected from the physical
world, but even more, cognition fundamentally relies on its relation to the
physical world (so-called \emph{embodied cognition}). Varela~\cite{Varela1992}
is one of the main discoverers of these mechanisms, and coined the concept of
\emph{enactivism} as the theoretical framework that studies the links between
cognition, embodiment and actions.  This has since been thoroughly studied in
robotics and artificial intelligence (Pfeifer and
Bongard~\cite{pfeifer2007body} is one of the reference).

The challenge of symbol grounding is also tightly linked to this issue. It
corresponds to the identification or creation, and then, maintenance of a link
between the symbol (the syntactic form of knowledge that the computer
manipulates) and its semantics (its meaning, anchored in the world). The
relation between the symbol, the referent of the symbol, and mediating minds
is classically referred as the \emph{semantic triangle} and has been
extensively studied in linguistics. The issue of grounding is well known in
cognitive science and is summarised by Harnad~\cite{Harnad1990} by this
question: ``how the semantic interpretation of a formal symbol system can be
made intrinsic to the system?''. This issue has a major practical importance in
robotic: for a robot to be both endowed with a symbolic representational and
reasoning system, and able to \emph{act} in the physical world, it must ground
its knowledge.

@@

Our context, where a robot has to achieve a task together with
humans, raises issues to be tackled by the robots' decisional components, and
that we can summarise as ``W-questions'': \emph{What}, \emph{Who}, \emph{When},
\emph{Where} and \emph{How}?

\emph{What to do next}, at different levels of abstractions and while taking
into account not only the current state but also the long term goals, is the
basic question for an intelligent robot. It is made more complex here by having
to deal with the partially observable physical and mental state of the human
partner, and by the extended set of possible actions.

@@

\fixme{Rachid: the reviewer 2 asks for a better connection between these general
questions and the specific approaches that we take}

@@

\fixme{Rachid -- TDB: what are the other 'deliberative architecture for socially interactive
robots' and where/how ours is better?}

\fixme{Discussion on deliberative architectures != cognitive architecture}: a \emph{cognitive
architecture} is usually understood as an artificial yet principled model of (human)
cognition, while we propose a practical approach to consistently combine a large
set of independent cognitive processes by bridging them through high-level semantics.
